---
title: "Week 2 assignment: Recorded steps"
author: "Joshua Solomon"
date: "September 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this assignment is to demonstrate use of R Markdown and knitr in presentation of data.

##Loading and processing the data
The data needs to be processed first.
```{r getandcleandata}
unzip("activity.zip")
activity <- read.csv("activity.csv")
activity$interval <- sprintf("%04d", activity$interval)
activity$datetime <- as.POSIXct(strptime(as.character(paste(activity$date, activity$interval)), "%Y-%m-%d %H%M", tz = "America/New_York"))
if("dplyr" %in% rownames(installed.packages()) == FALSE) {
    install.packages("dplyr")}
library(dplyr)
activity <- activity %>% mutate(datetime = datetime) %>% mutate(date = as.Date(as.character(date), "%Y-%m-%d"))
```

##What is mean total number of steps taken per day?
To answer this question, I grouped the data, plotted a histogram, and found the mean and median of the data.  Note that NA values were removed.  Also of note is that the distribution is nearly normal.
```{r median_mean}
steptotal <- activity %>% group_by(date) %>%
    summarize(sum = sum(steps))
if("ggplot2" %in% rownames(installed.packages()) == FALSE) {
    install.packages("ggplot2")}
library(ggplot2)
qplot(steptotal$sum, bins = 8, xlab = "Steps", ylab = "Days", 
    main = "Steps taken in a day", na.rm = TRUE)
mean_median <- as.data.frame(summarize(steptotal, 
    Mean = mean(sum, na.rm = TRUE), 
    Median = median(sum, na.rm = TRUE)))
print(mean_median)
```

##What is the average daily activity pattern?
Daily activity pattern was assessed by averaging data at each time interval over the entire data set (NA's were ignored) and plotting a time series.
```{r daily_pattern}
activity$interval_time <- strftime(strptime(activity$interval, "%H%M",
    tz = "America/New_York"), "%H:%M:%S")
intervalsteps <- activity %>% group_by(interval_time) %>%
    summarize(steps_interval = mean(steps, na.rm = TRUE))
intervalsteps$POSIX <- as.POSIXct(intervalsteps$interval_time, 
    format = "%H:%M:%S", tz = "America/New_York")
if("scales" %in% rownames(installed.packages()) == FALSE) {
    install.packages("scales")}
library(scales)
ggplot(intervalsteps, aes(POSIX, steps_interval)) + geom_line() +
    scale_x_datetime(labels = date_format("%H:%M", 
    tz = "America/New_York")) + labs(x = "Time", 
    y = "Steps per interval", 
    title = "Steps at each interval per day")
```

The interval with the maximum average of steps was computed below:
```{r max_interval}
subset(as.data.frame(intervalsteps), 
    steps_interval %in% max(steps_interval), 
    select = c(interval_time, steps_interval))
```

## Imputing missing values
There are a large number of NA values in this set.
```{r NA_percentage}
mean(is.na(activity$steps))
```

Thus, I am imputing missing data.  To do this, I am using the K-means clusering algorithm in the impute package from Bioconductor.
```{r impute}
if("impute" %in% rownames(installed.packages()) == FALSE) {
    source("https://bioconductor.org/biocLite.R")
    biocLite("impute")}
library(impute)
impute <- as.data.frame(impute.knn(as.matrix(cbind(activity$steps,
    as.integer(activity$interval))))$data)
steps_impute <- impute %>% select(V1) %>% 
    rename(steps_imputed = V1)
activity_impute <- cbind(steps_impute, activity[,-1])
```

A histogram including the imputed data is plotted below:
```{r imputed_histogram}
step_impute_total <- activity_impute %>% group_by(date) %>%
    summarize(sum = sum(steps_imputed))
qplot(step_impute_total$sum, bins = 8, xlab = "Steps", ylab = "Days", 
    main = "Imputed steps taken in a day")
```

As seen above, only a minor change was made in one of the bars of the histogram.  Similarly, a very small decrease in mean and median were calculated below.  Although I have not done an inferential analysis, I feel that the change is not statistically significant.
```{r mean_median_imputed}
mean_median_imputed <- as.data.frame(summarize(step_impute_total, 
    Mean = mean(sum), 
    Median = median(sum)))
print(mean_median_imputed)
```

## Are there differences in activity patterns between weekdays and weekends?
In order to answer this question, the data have to be first be grouped by day of the week, then by weekday and weekend.
```{r weekday_weekend_grouping}
activity_impute <- mutate(activity_impute, weekday = weekdays(datetime))
weekdaylist <- subset(activity_impute, 
    weekday == "Monday" | weekday == "Tuesday" | 
    weekday == "Wednesday" | weekday == "Thursday" | 
    weekday == "Friday")
weekdaylist$day <- "Weekday"
weekendlist <- subset(activity_impute, weekday == "Saturday" |
    weekday == "Sunday")
weekendlist$day <- "Weekend"
activity_weekday <- rbind(weekdaylist, weekendlist)
```

To assess the patterns, I plotted the data between the two facets of **weekday** and **weekend**.  It appears that there is no spike at 6:00 AM on the weekend and that the spike around 8:30 AM is not as high in the weekend facet as it is in the weekday facet.  This makes sense in that the spikes probably refer to waking and getting to work on a weekday.
```{r weekday_weekend}
interval_impute <- activity_weekday %>% 
    group_by(interval_time, day) %>%
    summarize(steps_interval = mean(steps_imputed))
interval_impute$POSIX <- as.POSIXct(interval_impute$interval_time, 
    format = "%H:%M:%S", tz = "America/New_York")
ggplot(interval_impute, aes(POSIX, steps_interval)) + geom_line() + 
    facet_grid(. ~ day) +
    scale_x_datetime(labels = date_format("%H:%M", 
    tz = "America/New_York")) + labs(x = "Time", 
    y = "Steps per interval", 
    title = "Steps at each interval per day")
```